Imperative Programming with Variational Effects

### 2018-07-18 paper/paper.tex
more inference rules

Of course, to arrive at the above result, we simply worked backwards from the non-variational
results. In order to take advantage of the benefits provided by sharing, we must instead develop a model
of variational execution that supports mutable state. We assume

@@

>>>>

Of course, to arrive at the above result, we simply worked backwards from the non-variational
results. In order to take advantage of the benefits provided by sharing, we must instead develop a model
of variational execution that supports mutable state.  As we will see, the applicative functor instance for
\prog{V} is no longer sufficient by itself, necessitating a different approach to how we variationalize programs
that use mutable state.

The central problem becomes obvious when we compare the required semantic domains for evaluating statements
in the plain and variational versions of the example. The semantic domain of plain statements is a function from state to
state, where state is a mapping from variable names to values. As such, we might then assume that the corresponding
semantic domain for variational statements is also from state to state, only in this case we use a variational map. But when
we attempt to evaluate our example under this semantic domain we soon reach an issue that causes us to rethink our definition.

The problem comes in line 4, where we assign the variable $x$ in a conditional block. In the previous line, we check if the value of
$y$ is true. When $y$ is a variational value, allowing it to be both true or false in different variants, we only assign $x$ in those variants
in which $y$ is true. But if the state in our semantic domain is only a map, we have no way of passing the necessary information to the
assignment that it must only occur in particular variants and not in others. Our only choice is to assign the value of $x$ for all variants,
which would clearly be incorrect, as the assignment should only be carried out in the variant where $y$ is true. We must therefore
conclude that we need to carry this variational context information around in our state during execution in order to handle cases where
assignment occurs in a branch that is conditionally executed. Similar logic can be used to see that this context is also necessary to handle
variational lookup of variable values.

It is now clear why our previous approach of relying solely on the applicative functor instance for choices is inadequate. The method provides
no way of maintaining and manipulating variational context during execution. This nicely demonstrates the problems we encounter when
trying to introduce choices to languages with side effects.

%\begin{program}
%type Var = String
%data AExpr = N Int | Ref Var | GetSecret
%data BExpr = Equ AExpr AExpr
%data Stmt = Assn Var AExpr | If BExpr Block
%type Block = [Stmt]
%
%evalAExpr :: AExpr -> State -> Int
%evalAExpr (N i) _ = i
%evalAExpr (Ref x) s = lookup x s
%evalAExpr GetSecret = undefined
%
%evalBExpr :: BExpr -> State -> Bool
%evalBExpr (Equ l r) s = evalAExpr l s == evalAExpr r s
%
%evalStmt :: Stmt -> State -> State
%evalStmt (Assn x e) s = update x (evalAExpr e) s
%evalStmt (If e ss) s | evalBExpr e s = evalBlock ss s
%                     | otherwise     = s
% 
%evalBlock :: Block -> State -> State
%evalBlock b s = foldl (flip evalStmt) s b
%\end{program}

@@

\newcommand{\GetCtx}[2]{\mathit{ctx}(#1,#2)}
\def \T {\mathit{true}}
\def \F {\mathit{false}}
\def \BigVErr {\infer [VS-Err] { } {(C,S,E,\T,s) \Downarrow_{VS} (S,E,\T)}}
\def \BigVSkip {\infer [VS-Skip] { } {(C,S,E,F,\text{skip}) \Downarrow_{VS} (S,E,F)}}
\def \BigVAssn {\infer [VS-Assn] {(\GetCtx{C}{E},S,e) \Downarrow_{VA} v} {(C,S,E,\F,x \coloneqq e) \Downarrow_{VS} (S[(\GetCtx{C}{E},x) \rightarrow v],E,\F)}}
\def \BigVSeq {\infer [VS-Seq] {(C,S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C,S',E',F,s') \Downarrow_{VS} (S'',E'',F')} {(C,S,E,s\ \text{seq}\ s') \Downarrow_{VS} (S'',E'',F')}}
\def \BigVIfOne {\infer [VS-If1] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(C \wedge \mathit{whenTrue}(v)) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ \mathit{sat}(C \wedge \mathit{whenFalse}(v)) \\ (C \wedge \mathit{whenFalse}(v),S',E',\F) \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S'',E'',F \wedge F')}}

\begin{mathpar}
\BigVErr \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIfOne
\end{mathpar}
